---
title: "Cod profile"
date: "07-06-2020"
output: html_notebook
---

I don't expect to see a big change in cod due to temperature--reports about it being more or less in the middle of its range. Changes might be more due to harvest regulations.

Beaugrand et al. 2008 puts ideal cod "regime" as high Chl, low diversity but large copepods, annual SST < 10. 9-10C marks boundary between Atlantic Polar and Atlantic Westerly Winds biomes--southern edge of cod distribution. Drinkwater 2005 puts range at -1-20, ideal 0-12.Expects recruitment to increase until 5C bottom temp, and decrease after 8.5 bottom temp.

MFRI advice report: https://www.hafogvatn.is/static/extras/images/01-cod1206941.pdf

```{r setup_data, eval = FALSE}
library(tidyverse)
library(gridExtra)
library (rfishbase)

# sample ID of all cod presences and absences
cod_pa <- read_csv ("../Data/PA_MFRI.csv",
                     col_types = cols(
                       sample_id = col_factor()
                     )) %>%
  dplyr::select (sample_id, Gadus_morhua)

# full MFRI info for all cod presences, retaining length data
cod_samp_full <- read_csv ("../Data/MFRI_comb_survey.csv",
                      col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor()
                      )
                      )%>%
  filter (species == 1, !(year == 2011 & season == "autumn")) # remove autumn 2011
  
# all cod presences, grouped by sample_ID (lengths aggregated
cod_samp <- cod_samp_full %>%
  group_by (sample_id) %>%
  summarize (n_tot = sum(n_per_nautmile, na.rm = TRUE),
             kg_tot = sum(kg_per_nautmile, na.rm = TRUE),
             station = first(station),
             lat = first(lat),
             lon = first(lon),
             year = first(year),
             date = first(date),
             season = first(season)
  ) %>%
  mutate (n_log = log (n_tot),
          kg_log = log(kg_tot))


# # predictor variables
# I compiled these in Build_MFRI_predictor_df.R. 
mfri_pred <-  read_csv ("../Data/MFRI_predictor_df.csv",
                        col_types = cols(
                          sample_id = col_factor(),
                          stat_sq = col_factor(),
                          bormicon_region = col_factor(),
                          sst_dev = col_number(),
                          bt_dev = col_number(),
                          sst_max = col_number(),
                          sst_min = col_number(),
                          bt_max = col_number(),
                          bt_min = col_number()
                        )
) %>%
  filter (!(year == 2011 & season == "autumn"),
          !(year < 2000 & season == "autumn")
  )  # remove autumn 2011, 131 samples. Also cutting out autumn pre-2000 (as of 7/23/2020. so did this with MASE fitting, but not with original run through of full gams)

```

### Fishbase stats
```{r fishbase}
species("Gadus morhua")

# temperature data
# https://github.com/ropensci/rfishbase/issues/47
stocks("Gadus morhua",  c("TempMin", "TempMax", "StockDefs")) #%>%
  filter (StockCode == 56739	)
```
Not a lot of info for iceland stock. temperature only there for north atlantic Canada/NA stock, 0-15 C
### Histograms
```{r}
cod_samp_full %>%
  ggplot (aes (y = length_cm, x = season), alpha = 0.5) +
  geom_violin()

cod_samp_full %>%
  ggplot (aes (x = length_cm, y = n_per_nautmile, fill = season)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.9) 

cod_samp %>%
  ggplot (aes (x = n_tot, fill = season), alpha = 0.2) +
  geom_histogram()
```

Fishbase puts age at maturity at 30-74cm. Quartiles for length are 37, 55 mean/median, 72, 170, so that actually works nicely. 

### Trends in abundance and biomass over time
```{r time_series, warning = FALSE}
biomass_ts <- cod_samp %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_kg, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Biomass annual mean")+
    theme(legend.position = "none")


ab_ts <- cod_samp %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_n, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Abundance annual mean")+
    theme(legend.position = "none")

grid.arrange (biomass_ts, ab_ts, nrow = 1)
```
### One possible outlier. 379179 sample ID
```{r}
cod_samp_full %>%
  filter (sample_id == 379179) %>% View()
```
## Spatial distribution

```{r plot_PA_points}

library (maps)
library (mapdata)
library (sf)
library (rnaturalearth)
library (rnaturalearthdata)
library (rgeos)
library (viridis)

world <- ne_countries(scale = "medium", returnclass = "sf")

 spp_PA <- cod_pa %>%
    left_join (mfri_pred, by = "sample_id") 

png (file = "../Figures/Cod_PA_points.png", width = 6, height = 4, units = "in", res = 300)
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  geom_point(data = spp_PA, pch = 21,
             aes (x = lon, y = lat, 
                  fill = as.factor(Gadus_morhua))) +
    scale_fill_manual (values = c ("white", "black"), name = "", guide = "none") +
    theme_bw() +
  labs(x = "", y = "") +
  theme(
    axis.text = element_blank(),
    plot.title = element_text (size = 24)
    #legend.text = element_text (size = 18),
    #legend.title = element_text (size = 20),
    #legend.position = c(0.9, 0.9)
  ) +
    ggtitle ("Groundfish survey tows")
dev.off()

```

```{r}

# https://trucvietle.me/r/tutorial/2017/01/18/spatial-heat-map-plotting-using-r.html

# this one seems closer:
# https://stackoverflow.com/questions/18285415/density2d-plot-using-another-variable-for-the-fill-similar-to-geom-tile
library (maps)
library (mapdata)
library (sf)
library (rnaturalearth)
library (rnaturalearthdata)
library (rgeos)
library (viridis)

world <- ne_countries(scale = "medium", returnclass = "sf")

  ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  geom_point(data = cod_samp, aes (x = lon, y = lat, col = n_tot)) +
    scale_fill_viridis_c()
  
  ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = cod_samp, aes (x = lon, y = lat, z = n_tot), fun = mean, binwidth = c (.25, .25)) +
    scale_fill_viridis_c(name = "Mean abundance") +
    theme_bw()
  
  ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = cod_samp, aes (x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.25, .25)) +
    scale_fill_viridis_c(name = "Mean abundance") +
    theme_bw() +
    facet_wrap (~season)
  
```
Look at spatial distribution by length class
```{r}
length_bins = quantile (cod_samp_full$length_cm)
length_bins[1] <- 0 # include smallest ones

cod_length_bins <- cod_samp_full %>%
  mutate (length_bin = cut(length_cm, breaks = length_bins),
          n_log = log(n_per_nautmile),
          n_tot = sum(n_per_nautmile))
  
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = cod_length_bins, aes(x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.5, .5)) +
    scale_fill_viridis_c(name = "Mean log abundance") +
    theme_bw() +
  theme (axis.text = element_blank()) +
    facet_grid (season ~ length_bin)

# facet grid for double facet 
# http://zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/#specify-a-grid-of-plots-by-row-and-column-with-facet_grid
  
```
not reallt enough resolution in the fall to tell if there's a difference. 

### spatial distribution over time
```{r}

all_yrs = data.frame (year = rep(1985:2020, each = 2),
                      season = rep (c("spring", "autumn"), length(1985:2020))
                      )

yr_breaks <- seq(1985, 2020, 5) 
yr_breaks[1] <- 0 # include 1985

cod_samp_yr_bn <- cod_samp %>%
      mutate (year_bn = cut(year, breaks = yr_breaks)) 

ggplot(data = world) +
      geom_sf() +
      coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
      stat_summary_2d(data = cod_samp_yr_bn, aes(x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.5, .5)) +
      scale_fill_viridis_c(name = "Mean log abundance") +
      facet_wrap (~year_bn) +
      theme_bw() +
      theme (axis.text = element_blank())

```
### Plot abundance vs. predictor variables
```{r cod_preds}
cod_ab_pred <- cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  right_join (mfri_pred, by = "sample_id")

  ggplot (cod_ab_pred, aes (x = bottom_temp, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = surface_temp, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = gins_sal, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = lat, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = lon, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = tow_depth_begin, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = tow_depth_end, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")

cod_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  left_join (mfri_pred, by = "sample_id") %>%
  ggplot (aes (x = oisst, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")


```
slightly negative rltp with surface and bottom temp. slightly positive lat and lon--makes sense, highest density in ne corner. slightly negative with depth but a little suspect since fewer deep tows. 

salinity for sample 358554 is unusually high. 
```{r}
summary (lm (kg_log ~ gins_sal, data = cod_ab_pred))

tmp_sal <- cod_ab_pred %>%
  filter (! sample_id == 358554) 
summary (lm (kg_log ~ gins_sal, data = tmp_sal))
```
doesn't change it all that much

satellite vs in situ temp
```{r}
summary (lm (kg_log ~ oisst, data = cod_ab_pred))
summary (lm (kg_log ~ surface_temp, data = cod_ab_pred))

summary (lm (kg_log ~ gins_bt, data = cod_ab_pred))
summary (lm (kg_log ~ bottom_temp, data = cod_ab_pred))
```



lil better with oisst and gins_bt. but many more missing observations for gins

```{r gam_results_explore}
load ("../Models/PresAbs/Gadus_morhua_PA_full.Rdata")
summary (gam_PA)

```
```{r gam_drop_sal}

cod_full_dropsal <- gam (Presence ~ te(lon, lat) + year + season + s(surface_temp) + s(bottom_temp) +s(tow_depth_begin),
                 family = "binomial", 
                 data = cod_pred)

summary (cod_full_dropsal)

```

