---
title: "Explore MFRI survey data"
date: "2020-06-16"
output: html_notebook
---

Pamela Woods provided autumn and spring survey data on 6/16/20. Performing preliminary data exploration here. 

```{r setup}
library(tidyverse)
library (skimr)
library (lubridate)

spr_ds <- read_csv ("../Data/springsurvey.csv",
                    col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor() # staves off trailing ch parsing error
      
                    )) %>%
  mutate (date = make_date (year, month))

aut_ds <- read_csv ("../Data/autumnsurvey.csv",
                    col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor() # staves off trailing ch parsing error
      
                    ))%>%
  mutate (date = make_date (year, month))

# combined dataset
comb_ds <- rbind (spr_ds, aut_ds) 
comb_ds$season = c(rep ("spring", nrow (spr_ds)), rep("autumn", nrow (aut_ds)))

# also create datasets with just station points, for temp and gear info
spr_samples <- spr_ds %>%
distinct_at ("sample_id", .keep_all = TRUE)

comb_samples <- aut_ds %>%
distinct_at ("sample_id", .keep_all = TRUE)

#combine
comb_samples <- rbind (spr_samples, aut_samples)
comb_samples$season = c(rep ("spring", nrow (spr_samples)), rep("autumn", nrow (aut_samples)))
```

```{r skim}
skim (aut_ds)
skim (spr_ds)
```

Spring survey has 20297 samples (feb-apr), 1985-2020. 85 unique spp, top are cod(ID 1), haddock(2), redfish (5), long rough dab (28). Autumn has 8705, 1995-2019 (sep-nov), 82 spp, same top 5. 29002 total samples, 465 stat_sq, 1094 stations. 

### Temporal extent
```{r tow_months}
comb_samples %>%
  group_by (month(date)) %>%
  summarize (count = n()) %>%
  rename (month = `month(date)`) %>%
  ggplot (aes (x = month, y = count)) +
  geom_bar(stat = 'identity') +
  geom_text (aes (label = count), vjust = 0.1)


unique (comb_ds$year[which(comb_ds$month ==4)])
unique (comb_ds$year[which(comb_ds$month ==)])
unique (comb_ds$year[which(comb_ds$month ==11)])

```
Only 33 samples in april, really negligible. Only in 1998 and 2005. 

```{r tow_years}
comb_samples %>%
  ggplot()+
  geom_bar(aes(x = year, fill = season), bins = 36) +
  ggtitle ("Tows per year")
```
doesn't look like any particular trend over time for spring. Autumn expanded sampling in 2001 and then fairly constant except for a very low year in 2011. 
#### Temporal extent of sampling stations
```{r station_duration}
#start with stat_Sq

#want a table with presence/absence of each stat sq, with years as columns?
comb_samples %>%
  group_by (stat_sq) %>%
  summarize (first_yr =) %>%
  ggplot (aes (x = year, y = factor(station))) +
  geom_point()

```

### Spatial extent
```{r tow_locations}

comb_samples %>%
  ggplot () +
  geom_boxplot(aes (x = season, y = lat))

comb_samples %>%
  ggplot () +
  geom_boxplot(aes (x = season, y = lon))

```
Lon looks the same, although autumn has a wider spread. Spring has a more northern median, autumn also wider spread. 
```{r tow_loc_time}
comb_samples %>%
  ggplot () +
  geom_violin(aes (x = factor(month(date)), y = lat))

comb_samples %>%
  ggplot () +
  geom_violin(aes (x = factor(month(date)), y = lon))
```

plot on map
```{r, spr_map}
library (ggmap)

isl_basemap <- get_map (location = c(-30, 60, -5, 70), source = "stamen", maptype = "terrain", crop = FALSE)

ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat), data = spr_samples) +
  ggtitle ("Spring survey locations")
```


```{r, fall_map}

ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat), data = aut_samples) +
  ggtitle ("Fall survey locations")

```

### Temperature data
```{r temp_hist}

ggplot (comb_samples) +
  geom_density (aes (x = surface_temp, col = season))

ggplot (comb_samples) +
  geom_boxplot (aes (x = season, y = surface_temp))

ggplot (comb_samples) +
  geom_density (aes (x = bottom_temp, col = season))

ggplot (comb_samples) +
  geom_boxplot (aes (x = season, y =bottom_temp))
```

```{r, temp_ts}
ggplot (comb_samples) +
  geom_point (aes (x = date, y =bottom_temp))

temp_month_ts <- comb_samples %>%
  group_by(date) %>%
  summarise (
    count = n(),
    mn_sst = mean(surface_temp, na.rm = TRUE),
    se_sst = sd (surface_temp, na.rm = TRUE)/sqrt(count),
    mn_bt = mean (bottom_temp, na.rm = TRUE),
    se_bt = sd (bottom_temp, na.rm = TRUE)/sqrt(count)
  ) %>%
  mutate (season = ifelse (month(date) %in% c(2:4), "spring", "fall"))

sst_fall_fit <- lm (surface_temp ~ date, data = aut_samples)
summary (sst_fall_fit) # y = 5.05 + 1.6e-04x, adj 42 = 0.035

sst_spr_fit <- lm (surface_temp ~ date, data = spr_samples)
summary (sst_spr_fit)

temp_month_ts %>%
  ggplot (aes (x = date, y = mn_sst, col = season)) +
  geom_point() +
  geom_line() +
  geom_smooth (method = "lm")

temp_month_ts %>%
  ggplot (aes (x = date, y = mn_bt, col = season)) +
  geom_point() +
  geom_line() +
  geom_smooth (method = "lm")
```

```{r sst_map}
library (viridis)
ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat, col = surface_temp), data = comb_samples) +
  scale_color_viridis(option ="inferno") +
  ggtitle ("SST at all survey locations")
```

### Gear variables
```{r gear_boxplots}
comb_samples %>%
  pivot_longer (
    cols = starts_with("tow"),
    names_to = "tow_info",
    values_to = "value"
  ) %>%
ggplot () +
  geom_histogram(aes (x = value, fill = season)) +
  facet_wrap (~tow_info, scales = "free")
```
Tow depth begin and end are left-skewed but that makes sense. Tow direction is a bit all over but probably fine. Tow-length has some weird patterns, there seems to be a few standard lengths and then some scatter. Tow speed and time look okay? Potentially different protocols for autumn and spring? Different depths, and peaks for length and time. 
```{r gear_dotcharts}
dotchart(comb_samples$tow_time, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow time")

dotchart(comb_samples$tow_length, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow length")

dotchart(comb_samples$tow_depth_begin, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow depth, start")
```
Autumn samples are much deeper. 

```{r gear_depth_rltps}
comb_samples %>%
  ggplot (aes (x = tow_depth_begin, y = tow_length, col = season)) +
  geom_point() +
  ggtitle("Tow length vs tow depth")

comb_samples %>%
  ggplot (aes (x = tow_depth_begin, y = tow_time, col = season)) +
  geom_point() +
  ggtitle ("tow time vs tow depth")
```
Tow times are fairly standard. The longest one is the deepest and the shortest ones are fairly shallow, but there are some long shallow tows too. 


### Species info

quick check about length biomass etc, might want to animate
```{r}
# cod_kg_ts <- comb_samples %>%
#   filter (species == 1) %>%
#   group_by (season, year(date), length_cm) %>%
#   summarise (mn_kg = mean (kg_per_nautmile, na.rm = TRUE))
#   
# ggplot (cod_kg_ts,
#         aes (x = `year(date)`, y = mn_kg, col = as.factor(length_cm))) +
#   geom_point() + 
#   geom_line() +
#     facet_wrap (~season)

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = date, y = kg_per_nautmile)) +
  geom_point()

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = lat, y = kg_per_nautmile)) +
  geom_point

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = lon, y = kg_per_nautmile)) +
  geom_point()

```

Which species can I cut out? any that have fewer than 10 in any month?
```{r spp_tally}
#tally samples 
spp_tally <- comb_ds %>%
  group_by (species) %>%
  tally() %>%
  arrange(desc(n))# %>%
  #View()


  summarise (count = n(), 
             mn_kg = mean (kg_per_nautmile, na.rm = TRUE), 
             mn_n = mean(n_per_nautmile, na.rm = TRUE)
             ) %>%
  filter (count > 0) %>%
  View()
  
```
91 total species. 6 with only one entry: 73 (Ciliata mustela fivebearded rockling), 46 (Iceland cyprine), 91 (lycodes rossi, threespot eelpout), 77 (mussel), 95 (psetta maxima turbot), 42 (spider crab)
