---
title: "Explore MFRI survey data"
date: "2020-06-16"
output: html_notebook
---

Pamela Woods provided autumn and spring survey data on 6/16/20. Performing preliminary data exploration here. 

Info on the survey, from 2020 survey manual:
Covers the shelf to depths of 500m. 
Stratification and allocation of stations is divided into Northern and Southern regions. Allocation based on pre-estimated cod density patterns. 

Bottom trawl, 3.8knots for 4nm. several tows are shorter than 4nm? 2nm min. 

```{r setup}
library(tidyverse)
library (skimr)
library (lubridate)

# combined autumn and spring survey data. Each row is a length observation of a species. 
comb_ds <- read_csv ("../Data/MFRI_comb_survey.csv",
                      col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor())
)

# combined autumn and spring survey data, with each sample tow as a row
comb_samples <- read_csv("../Data/MFRI_comb_survey_samples.csv",
                          col_types = cols(
                      sample_id = col_factor(),
                      stat_sq = col_factor())
)
```

```{r skim}

# quick check of data contents
skim (comb_ds)

```

Spring survey has 20297 samples (feb-apr), 1985-2020. 85 unique spp, top are cod(ID 1), haddock(2), redfish (5), long rough dab (28). Autumn has 8705, 1995-2019 (sep-nov), 82 spp, same top 5. 29002 total samples, 465 stat_sq, 1094 stations. 

Briefly panicked because comb_ds has 91 unique species but comb_samples has only 79. This is expected; just took distinct samples, not distinct samples and species. In my full PA set, have full 91 species. 

Discovered that sebasetes mentella (61, 11) is repeated. Majority of samples are 61; only 16 have 11. Will ask, but replace 11 with 61 for now. Magnus confirmed that this is a separate subspecies, so shouldn't combine. 

### Making sense of station, stat_sq, sample ID
Pamela: The maps show the target tow locations (can vary a little from year to year), which may very slightly, but tow numbers overlap so the 7-digit identifier is calculated as: statisticalsquare*10000 + townumber*100 + gear to make sure that they are unique to each spot on the map.

```{r station_sq_sample}
length (unique (comb_ds$sample_id))
length (unique (comb_ds$station))
length (unique (comb_ds$stat_sq))

comb_samples %>%
  group_by (season) %>%
  summarize (n_station = length (unique (station)),
             n_stat_sq = length (unique (stat_sq)))

comb_samples %>%
  group_by (season, year) %>%
  summarize (n_sample = length (unique (sample_id)),
             n_station = length (unique (station)),
             n_stat_sq = length (unique (stat_sq)))
```
29002 total sample_ID, 1094 station, 465 stat_sq. Autum: 414 stations, 340 squares. Spring, 680 stations, 336 stat sq. Same number of samples and stations each year. 

Over how many years were the stat_sqs sampled? Pamela mentioned that autumn is only consistently sampled after 2000, so might expect some stat_sqs in the 1990s that are only sampled a few times. 
```{r stat_sq_yrs_sampled}

single_yr_sq <- comb_samples %>%
  filter (season == "autumn", year != 2011) %>%
  group_by (stat_sq) %>%
  summarize (n_years = length (unique (year))) %>%
  filter (n_years < 5)

comb_samples %>%
  filter (season == "autumn", stat_sq %in% single_yr_sq$stat_sq) %>%
  dplyr:: select (stat_sq, year) %>%
  View()

```

Actually quite a few stat_sqs only sampled once throughout the 2000s...

### Temporal extent

Pamela: generally we only use autumn survey data from 2000 onwards to calculate survey indices due to station changes


```{r tow_months}
comb_samples %>%
  group_by (month(date)) %>%
  summarize (count = n()) %>%
  rename (month = `month(date)`) %>%
  ggplot (aes (x = month, y = count)) +
  geom_bar(stat = 'identity') +
  geom_text (aes (label = count), vjust = 0.1)


unique (comb_ds$year[which(comb_ds$month ==4)])
unique (comb_ds$year[which(comb_ds$month == 2)])
unique (comb_ds$year[which(comb_ds$month ==11)])

```
Only 33 samples in april, really negligible. Only in 1998 and 2005. 

```{r tow_years}
comb_samples %>%
  ggplot()+
  geom_bar(aes(x = year, fill = season), bins = 36) +
  ggtitle ("Tows per year")

# n samples each year
comb_samples %>%
  group_by (season, year) %>%
  summarize (count = n())

# avg samples each year, not counting 2011
comb_samples %>%
  filter (season == "autumn", year >= 2000, year != 2011) %>%
  group_by (year) %>%
  summarize (count = n()) %>%
  summarize (mean_samp = mean (count))
```
doesn't look like any particular trend over time for spring. Autumn expanded sampling in 2001 and then fairly constant except for a very low year in 2011. Pamela and Magnus explained that there was a labor strike in 2011, suggested just omitting that year. 
```{r autumn_2011}

comb_samples %>%
  filter (year == 2011, season == "autumn") %>%
  nrow()

comb_samples %>%
  filter (year == 2011, season == "autumn") %>%
  summarize (n_stat_sq = length (unique (stat_sq)),
             n_station = length (unique (station)))

```
131 tows in autumn 2011, vs mean 380 in other years since 2000.

#### Temporal extent of sampling stations
I basically want a presence/absence table, but with stations. There should be around 590 stations
```{r station_duration}

all_years <- data.frame (year = 1985:2020)

#start with stat_Sq
# separate fall and spring?
stat_sq_ts_spr <- comb_samples %>%
  filter (season == "spring") %>%
  distinct_at (c("stat_sq", "year")) %>%
  group_by (stat_sq) %>%
  summarize (count = n())

length (which (stat_sq_ts_spr$count == 36)) # 205
length (which (stat_sq_ts_spr$count > 30)) # 255
length (which (stat_sq_ts_spr$count < 5)) # 43 less than 5, 48 < 10

stat_sq_ts_aut <- comb_samples %>%
  filter (season == "autumn") %>%
  distinct_at (c("stat_sq", "year")) %>%
  group_by (stat_sq) %>%
  summarize (count = n())  # 340

length (stat_sq_ts_aut$stat_sq[which (stat_sq_ts_aut$stat_sq %in% stat_sq_ts_spr$stat_sq)]) # 211 shared

length (which (stat_sq_ts_aut$count == 25))
length (which (stat_sq_ts_aut$count > 24))

length (stat_sq_ts_aut$stat_sq[which (stat_sq_ts_aut$stat_sq %in% stat_sq_ts_spr$stat_sq & stat_sq_ts_aut$count > 20)]) # 161

#want a table with presence/absence of each stat sq, with years as columns?
station_ts_spr <- comb_samples %>%
  filter (season == "spring") %>%
  distinct_at (c("station", "year")) %>%
  group_by (station) %>%
  summarize (count = n())

hist (station_ts_spr$count)
length (which (station_ts_spr$count ==36)) # 436
length (which (station_ts_spr$count > 30)) # 518

station_ts_aut <- comb_samples %>%
  filter (season == "autumn") %>%
  distinct_at (c("station", "year")) %>%
  group_by (station) %>%
  summarize (count = n())

hist (station_ts_aut$count)

length (which (station_ts_aut$count >20)) # 280

length (which (station_ts_aut$station %in% station_ts_spr$station))

comb_samples %>%
  group_by(year, season) %>%
  summarise (n_stations = length (unique(station))) %>%
  ggplot (aes (x = year, y = n_stations, col = season)) +
  geom_point() +
  geom_line()

```
336 stat_sq in spring; 205 sampled every year, 255 sampled in 30+ years. 340 stat_sq in autumn, 211 shared with spring. Only 34 sampled all years. 230 > 20 years. 211>22, 150>23. 

680 stations in spring; 436 in all years, 518 > 30. 414 in autumn, 280 above 20. No overlap between spring and fall.

From Pamela: The maps show the target tow locations (can vary a little from year to year), which may very slightly, but tow numbers overlap so the 7-digit identifier is calculated as: statisticalsquare*10000 + townumber*100 + gear to make sure that they are unique to each spot on the map.
```{r gear}

length (unique (comb_samples$gear)) #3
comb_samples %>%
  group_by (season, gear) %>%
  summarize (count = n())

comb_samples %>%
  group_by (season, stat_sq) %>%
  summarize (count = n())

comb_samples %>%
  filter (season == "autumn") %>%
  group_by (year, gear) %>% 
  summarize (count = n()) 

comb_samples %>%
  filter (season == "autumn") %>%
  group_by (stat_sq, year) %>% 
  summarize (count = n()) %>% View()

```
Spring is all 73. Autumn is half 77, 78, seems like they do about half and half each year except only 77 in 1995. Small number of both in 2011. 

### Spatial extent
```{r tow_locations}

comb_samples %>%
  ggplot () +
  geom_boxplot(aes (x = season, y = lat))

comb_samples %>%
  ggplot () +
  geom_boxplot(aes (x = season, y = lon))

```
Lon looks the same, although autumn has a wider spread. Spring has a more northern median, autumn also wider spread. 
```{r tow_loc_time}
comb_samples %>%
  ggplot () +
  geom_violin(aes (x = factor(month(date)), y = lat))

comb_samples %>%
  ggplot () +
  geom_violin(aes (x = factor(month(date)), y = lon))
```

plot on map
```{r spr_map}
library (ggmap)

isl_basemap <- get_map (location = c(-30, 60, -5, 70), source = "stamen", maptype = "terrain", crop = FALSE)

ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat), data = spr_samples) +
  ggtitle ("Spring survey locations")
```


```{r fall_map}

ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat), data = aut_samples) +
  ggtitle ("Fall survey locations")

```

```{r tow_depth}
comb_samples %>% ggplot (aes (x = season, y = tow_depth_begin)) + geom_boxplot()
comb_samples %>% ggplot (aes (x = season, y = tow_depth_end)) + geom_boxplot()

comb_samples %>% ggplot (aes (x = tow_depth_begin, col = season)) + geom_density()

comb_samples %>% ggplot (aes (x = tow_depth_end, col = season)) + geom_density()
```
Autumn has deeper samples, but just a much fatter tail. still has mostly shallower samples. Spring also has a few deep ones. 

### Temperature data

How extensive is temperature sampling?
```{r n_temp_samples}

length (which (is.na (comb_samples$surface_temp))) # 953 missing values;

length (which (is.na (comb_samples$bottom_temp))) # 1240

length (which (is.na (comb_samples$surface_temp) & is.na (comb_samples$bottom_temp))) # 151

comb_samples %>%
  filter (!is.na (surface_temp) & !is.na(bottom_temp)) %>%
  nrow()

```
953 missing surface temp (leaves 28049), 1240 missing bottom temp (leaves 27762). 151 missing both, so not actually very much overlap--lose 2042 total. 26,960 have both surface and bottom. 
```{r temp_hist}

ggplot (comb_samples) +
  geom_density (aes (x = surface_temp, col = season))

ggplot (comb_samples) +
  geom_boxplot (aes (x = season, y = surface_temp))

ggplot (comb_samples) +
  geom_density (aes (x = bottom_temp, col = season))

ggplot (comb_samples) +
  geom_boxplot (aes (x = season, y =bottom_temp))
```
SST: Warmer temps in the fall, peak around 9 degrees and another small peak at 6. Spring peak at 3 and 4-7. Autumn has more spread, some values go below spring it looks like. 

Bottom temp: fall is bimodal,peak just below zero and at 7. Spring peak around 3 and 5. fall still warmer, with more spread. 
```{r temp_ts}
ggplot (comb_samples) +
  geom_point (aes (x = date, y =bottom_temp))

temp_month_ts <- comb_samples %>%
  group_by(date) %>%
  summarise (
    count = n(),
    mn_sst = mean(surface_temp, na.rm = TRUE),
    se_sst = sd (surface_temp, na.rm = TRUE)/sqrt(count),
    mn_bt = mean (bottom_temp, na.rm = TRUE),
    se_bt = sd (bottom_temp, na.rm = TRUE)/sqrt(count)
  ) %>%
  mutate (season = ifelse (month(date) %in% c(2:4), "spring", "fall"))

sst_fall_fit <- lm (surface_temp ~ date, data = filter (comb_samples, season == "autumn"))
summary (sst_fall_fit) # y = 5.05 + 1.6e-04x, adj R2 = 0.035

sst_spr_fit <- lm (surface_temp ~ date, data = filter (comb_samples, season == "spring"))
summary (sst_spr_fit) # y = 3.27 + 9.67e-05x, adj R2 = 0.028

temp_month_ts %>%
  ggplot (aes (x = date, y = mn_sst, col = season)) +
  geom_point() +
  geom_line() +
  geom_smooth (method = "lm")

temp_month_ts %>%
  ggplot (aes (x = date, y = mn_bt, col = season)) +
  geom_point() +
  geom_line() +
  geom_smooth (method = "lm")
```
SST increasing in both seasons. BT too, much more variable. 
```{r sst_map}
library (viridis)
ggmap(isl_basemap) +
  geom_point (aes (x = lon, y = lat, col = surface_temp), data = comb_samples) +
  scale_color_viridis(option ="inferno") +
  ggtitle ("SST at all survey locations")
```

```{r temp_lat}
ggplot (comb_samples, aes (x = lat, y = surface_temp, col = season)) +
  geom_point(alpha= 0.5) +
  geom_smooth (method ="lm")

sst_fall_lat_fit <- lm (surface_temp ~ lat, data = filter (comb_samples, season == "autumn"))
summary (sst_fall_lat_fit) # y = 71 -.98x, adj R2 = 0.43

sst_spr_lat_fit <- lm (surface_temp ~ lat, data = filter (comb_samples, season == "spring"))
summary (sst_spr_lat_fit) # y = 81.55 -1.18x, adj R2 = 0.45
```

### Gear variables
```{r gear_boxplots}
comb_samples %>%
  pivot_longer (
    cols = starts_with("tow"),
    names_to = "tow_info",
    values_to = "value"
  ) %>%
ggplot () +
  geom_histogram(aes (x = value, fill = season)) +
  facet_wrap (~tow_info, scales = "free")
```
Tow depth begin and end are left-skewed but that makes sense. Tow direction is a bit all over but probably fine. Tow-length has some weird patterns, there seems to be a few standard lengths and then some scatter. Tow speed and time look okay? Potentially different protocols for autumn and spring? Different depths, and peaks for length and time. 
```{r gear_dotcharts}
dotchart(comb_samples$tow_time, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow time")

dotchart(comb_samples$tow_length, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow length")

dotchart(comb_samples$tow_depth_begin, 
         groups = factor(comb_samples$season), 
         pch = comb_samples$season,
         main = "Tow depth, start")
```
Autumn samples are much deeper. 

```{r gear_depth_rltps}
comb_samples %>%
  ggplot (aes (x = tow_depth_begin, y = tow_length, col = season)) +
  geom_point() +
  ggtitle("Tow length vs tow depth")

comb_samples %>%
  ggplot (aes (x = tow_depth_begin, y = tow_time, col = season)) +
  geom_point() +
  ggtitle ("tow time vs tow depth")
```
Tow times are fairly standard. The longest one is the deepest and the shortest ones are fairly shallow, but there are some long shallow tows too. 


### Species info

quick check about length biomass etc, might want to animate
```{r biomass_check}
# cod_kg_ts <- comb_samples %>%
#   filter (species == 1) %>%
#   group_by (season, year(date), length_cm) %>%
#   summarise (mn_kg = mean (kg_per_nautmile, na.rm = TRUE))
#   
# ggplot (cod_kg_ts,
#         aes (x = `year(date)`, y = mn_kg, col = as.factor(length_cm))) +
#   geom_point() + 
#   geom_line() +
#     facet_wrap (~season)

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = date, y = kg_per_nautmile)) +
  geom_point()

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = lat, y = kg_per_nautmile)) +
  geom_point

comb_samples %>%
  filter (species ==1) %>%
  ggplot (aes (x = lon, y = kg_per_nautmile)) +
  geom_point()

```

Which species can I cut out? any that have fewer than 10 in any month?
```{r spp_tally}
#tally samples 
spp_tally <- comb_ds %>%
  distinct_at (c("sample_id", "species"), .keep_all = TRUE) %>%
  group_by (species) %>%
  tally() %>%
  arrange(desc(n))# %>%
  #View()

hist (spp_tally$n, breaks = 30)

  
  spp_tally_yr <- comb_ds %>%
  distinct_at (c("sample_id", "species"), .keep_all = TRUE) %>%
  group_by (species, year) %>% 
    tally() 

  # filter 28 top spp
  major_spp <- filter (spp_list, Spp_ID %in% spp_tally_yr$species)
  
```
91 total species. 6 with only one entry: 73 (Ciliata mustela fivebearded rockling), 46 (Iceland cyprine), 91 (lycodes rossi, threespot eelpout), 77 (mussel), 95 (psetta maxima turbot), 42 (spider crab). Filtering for n>10 each year(wouldn't count no entries), get 76. n >100 each year, 60. n>500, 39. n>1000, 28.

11 spp have more than 10k samples.23 have more than 5000. 49 have more than 1000. 14 have fewer than 10, 

28 major spp: doesn't include monk or mackerel. 39 does have monkfish. 

Looking through abundance TS:
species with 2005-10 peak:
lemon sole, haddock, lumpfish(autumn)

downward trend:
starry skate

weird early 90s point, high blue whiting, arctic rockling

Look at PA full colsums
```{r PA_full_colsums}
mfri_pa <- read_csv ("../Data/PA_MFRI.csv",
                     col_types = cols(
                       sample_id = col_factor()
                     ))

# this has 90 spp

length (which (colSums (mfri_pa) > 5000)) # 24. 50 > 1000, 58 > 500

sample_yrs <- comb_samples %>%
  dplyr::select (sample_id, year, season)

spp_pa_yr <- mfri_pa %>%
  pivot_longer (-sample_id, 
                names_to = "species",
                values_to = "presence") %>%
  left_join (sample_yrs, by = "sample_id") %>%
  group_by (species, year, season) %>%
  summarize (n_p = sum(presence))

```

How many species have samples every year?
36 total years for spring, 25 total years for autumn. 
```{r sampled_all_yrs}
spp_yrs_sampled <- spp_pa_yr %>%
  mutate (p = ifelse (n_p > 0, 1, 0)) %>%
  group_by (species, season) %>%
  summarize (n_yrs_sampled = sum(p))

length(which (spp_yrs_sampled$season =="spring" & spp_yrs_sampled$n_yrs_sampled > 25)) 
length(which (spp_yrs_sampled$season =="autumn" & spp_yrs_sampled$n_yrs_sampled > 14)) 

# spring species consistently sampled
spring_spp <- unique(spp_yrs_sampled$species[which (spp_yrs_sampled$season =="spring" & spp_yrs_sampled$n_yrs_sampled ==36)]) 

# fall spp consistently sampled
fall_spp <- unique(spp_yrs_sampled$species[which (spp_yrs_sampled$season =="autumn" & spp_yrs_sampled$n_yrs_sampled > 24)])

#overlap in fall/spring consistent spp
fall_spp[which (fall_spp %in% spring_spp)]

spp_25 <- 

```
27 spp sampled every spring, 29 every autumn. 25 in both. 37 have more than 30, 31 have more than 32 in spring. 

52 have < 10 years missing in spring, 63 in fall. 

Pamela and Magnus suggested on 6/29 that there might be some species for which it doesn't make sense to sample both fall and spring. Particularly if there's a big difference in length data. plotting those differences in the plot_spp_distribution script. 

#### Check anomalous points
1990, big spike in blue whiting. Sample 44366 has unusually high numbers, just of blue whiting it seems. 
```{r blue_whiting_1990}

comb_ds %>% 
  filter (species == 34 & year == 1990) %>%
  View()

comb_ds %>% 
  filter (sample_id == 44366) %>%
  group_by (species) %>%
  summarize (tot_kg = sum(kg_per_nautmile)) %>%
  ggplot (aes (y = tot_kg, x = species)) +
  geom_boxplot()
  View()

```
mid-90s rabbitfish 39. Looks like 73586 in 1996. no particular values jump out, just a high abundance of many size classes.
```{r rabbitfish}
comb_ds %>% 
  filter (species == 39 & between(year, 1995, 1998) & kg_per_nautmile > 1) %>%
  View()
```

##### repeated lengths
noticed 7/8/20 several repeat lengths for cod. How would I best find which samples have repeated lengths? nrow (distinct) by sample ID and species and length vs nrow?
```{r}
check_length_repeats <- comb_ds %>%
  distinct_at (c("sample_id", "species", "length_cm"))

nrow (check_length_repeats) # 3597803


tmp <- comb_ds %>%
  group_by (sample_id, species, length_cm) %>%
  summarize (count = n()) %>%
  filter (count > 2) #%>%
  #summarize (samples = length (unique(sample_id)))
  

#nrow() #67926


```

#### Match landed and surveyed species
```{r landings_spp}
ldgs <- read_csv ("../Data/Landings_annual_all_sectors.csv")
spp_list <- read_csv("../Data/species_eng.csv")
match_spp <- read_csv("../Data/Match_spp_names.csv") %>% # this is only the 41 landed species. might be better to just add a column to spp_list?
  mutate (sci_name_underscore = gsub(" ", "_", Scientific_name)) # add underscore to match PA table

ld_spp <- sort (unique (ldgs$species))

ldgs %>%
  group_by (species) %>%
  summarize (tot_cat = sum(tonnes, na.rm = TRUE)) %>%
  arrange (desc (tot_cat))
```
41 landed species. Are any very rare in survey?
```{r landed_spp_n_yrs}
# just number of samples each year
spp_yrs_sampled %>%
  filter (species %in% match_spp$sci_name_underscore) %>%
  filter (n_yrs_sampled < 10) %>%
  View()

spp_yrs_sampled %>%
  filter (species %in% match_spp$sci_name_underscore) %>%
  #filter (n_yrs_sampled < 10) %>%
  View()
  
comb_ds %>%
  filter (species %in% match_spp$Spp_ID) %>%
  group_by (species, year) %>%
  summarize (tot_biomass = sum(kg_per_nautmile, na.rm = TRUE))

```
Grenadier C. rupestris only sampled in the autumn, but present in 24 years. Norway lobster, shrimp, and greenland shark don't have many samples. Mackerel also not well represented (14 autumn; 11 spring)

Are there commonly caught species that are NOT landed?
```{r not_landed}
spring_spp[which (!spring_spp %in% match_spp$sci_name_underscore)]
fall_spp[which (!fall_spp %in% match_spp$sci_name_underscore)]
```
Spring: A denticulatus (Arctic wolffish), lumpfish, Norway haddock [perhaps in with haddock or redfish?], Norway pout
Fall: Polar cod, lumpfish, gray gurnard, Norway haddock, Norway pout

Relationships btw species
```{r cod_ts_friends}

24
mfri_pa %>%
  dplyr::select (sample_id)

```


# to get actual sample distributions for each species, would need to add the zeros back in for the samples with no catch. 
```{r test_distributions}
#https://aosmith.rbind.io/2019/03/06/lots-of-zeros/

library(HMMpa)
library (MASS)
# 
# cod_bm <- comb_ds %>%
#   filter (species == 1)

```
### relationships btw predictor variables

```{r survey_var_pairplots}
library(lattice)
spatial_vars <- comb_samples %>%
  mutate(season_num = ifelse (season == "spring", 1, 2)) %>%
  dplyr::select (lat, lon, year, tow_depth_begin, surface_temp, bottom_temp, season_num)

# from the documentation...
## put histograms on the diagonal
panel.hist <- function(x, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(usr[1:2], 0, 1.5) )
    h <- hist(x, plot = FALSE)
    breaks <- h$breaks; nB <- length(breaks)
    y <- h$counts; y <- y/max(y)
    rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}


## put (absolute) correlations on the upper panels,
## with size proportional to the correlations.
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y, use = "na.or.complete"))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

pairs (spatial_vars, upper.panel = panel.cor, diag.panel = panel.hist)
```
See a 0.7 correlation between surface and bottom temp (pos), this makes sense. Also a negative relationship between lat and surface temp and bottom temp, this also makes sense--colder at higher latitudes. Would expect more of a relationship with depth and bottom temp. Year and depth makes sense, probably deeper corresponding with starting autumn surveys. 
```{r salinity}
mfri_pred <- read_csv ("../Data/MFRI_predictor_df.csv",
                       col_types = cols(
                      sample_id = col_factor(),

                      stat_sq = col_factor())
)


mfri_pred_vars <- mfri_pred %>%
  dplyr::select (year, month, lat, lon, oisst, gins_bt, gins_sal, bottom_temp, surface_temp)

pairs (mfri_pred_vars, upper.panel = panel.cor, diag.panel = panel.hist)

# just look at satellite vs in situ 
mfri_env <- mfri_pred %>%
  dplyr::select(oisst, gins_bt, gins_sal, bottom_temp, surface_temp) 
pairs (mfri_env, upper.panel = panel.cor, diag.panel = panel.hist)

# just look at the variables I put in my gam
gam_env <- mfri_pred %>%
  mutate(season_num = ifelse (season == "spring", 1, 2)) %>%
  dplyr::select(lat, lon, year, season_num, surface_temp, bottom_temp, gins_sal, tow_depth_begin)

png (filename = "../Figures/JM_GAM_pairplot.png", width = 11, height = 8.5, unit = "in", res = 72)
pairs (gam_env, upper.panel = panel.cor, diag.panel = panel.hist,
       main = "Pairplot, environmental variables used in full GAM")
dev.off()
```
Potential positive relationship between bottom time and salinity? 

Actually missing quite a bit of data. 8042 points for GINS, 353 for oisst. For GINS, 7261 points are after 2012. The remaining missing points are too close to the coast and the GINS data don't register. 

```{r corrplot_gins}
library(ggcorrplot)

mfri_comp <- mfri_pred_vars [ complete.cases (mfri_pred_vars),]
cont.corr <- round (cor (mfri_comp), 1)

ggcorrplot (cont.corr, hc.order = TRUE,
            type = "lower",
            lab = TRUE, 
            lab_size = 3,
            method = "circle",
            ggtheme = theme_bw
          ) 

```

Worrisome correlation between sst, bt, salinity, and lat. also month, salinity, bt and ssst. also salinity and bottom temp. 

```{r corrplot_mfri}
mfri_env <- mfri_pred %>%
  dplyr::select(bottom_temp, surface_temp, lat, lon, year, month) 

mfri_comp <- mfri_env [ complete.cases (mfri_env),]
cont.corr <- round (cor (mfri_comp), 1)

ggcorrplot (cont.corr, hc.order = TRUE,
            type = "lower",
            lab = TRUE, 
            lab_size = 3,
            method = "circle",
            ggtheme = theme_bw
          ) 

```


what is correlation between mfri temp and climatology temp
```{r mfri_gins_corrplot}
mfri_pred_vars <- mfri_pred %>%
  dplyr::select (year, month, lat, lon, oisst, gins_bt, gins_sal, surface_temp, bottom_temp)

mfri_comp <- mfri_pred_vars [ complete.cases (mfri_pred_vars),]
cont.corr <- round (cor (mfri_comp), 1)

ggcorrplot (cont.corr, hc.order = TRUE,
            type = "lower",
            lab = TRUE, 
            lab_size = 3,
            method = "circle",
            ggtheme = theme_bw
          )

```
surface temp and oisst very well correlated. bottom temp not as much, but still pretty good. 


### Corrplot with GLORYS temperature data
9/14/2020
```{r glorys_corrplot}
library(ggcorrplot)

mfri_pred <-  read_csv ("../Data/MFRI_predictor_df.csv",
                       col_types = cols(
                      sample_id = col_factor(),
                      stat_sq = col_factor(),
                      bormicon_region = col_factor(),
                      sst_dev = col_number(),
                      bt_dev = col_number(),
                      sst_max = col_number(),
                      sst_min = col_number(),
                      bt_max = col_number(),
                      bt_min = col_number())
)

mfri_pred_vars <- mfri_pred %>%
  dplyr::select ( tow_depth_begin, surface_temp, bottom_temp, sst_dev, bt_dev, sst_max, sst_min, bt_max) #, bt_min)

mfri_comp <- mfri_pred_vars [ complete.cases (mfri_pred_vars),]
cont.corr <- round (cor (mfri_comp), 1)

ggcorrplot (cont.corr, hc.order = TRUE,
            type = "lower",
            lab = TRUE, 
            lab_size = 3,
            method = "circle",
            ggtheme = theme_bw
          )

```
-0.8 with sst_min and lat. 0.8 with bottom_Temp and bt_min. 0.7 for bt_max and bt, bt_max/bt_min. 0.7 for sst and sst_min, sst_max and sst_min, sst_min and bt_min. -0.7 sst_max and lat. 

I think if I take out sst_min and bt_min or bottom temp I'd be in pretty good shape. might lean toward taking out bottom temp, bt_min is looking more helpful. 
sst_min looks more important than sst_max or sst however, although maybe should just try AIC

Looking at temperature_AIC_check.csv, dropping bt is much more problematic than dropping bt_min. 

# bormicon region and depth, temp
```{r bormicon}

mfri_pred %>%
  ggplot (aes (x = bormicon_region, y = tow_depth_begin)) +
  geom_boxplot()

mfri_pred %>%
  ggplot (aes (x = bormicon_region, y = surface_temp)) +
  geom_boxplot()

mfri_pred %>%
  ggplot (aes (x = bormicon_region, y = bottom_temp)) +
  geom_boxplot()


```
Definitely big differences in depth. 3 in particular are below 1000
surface temp also some differences, but actually fairly similar spread. bottom temp bigger differences, where deeper sites prob have lower temps. 



## Run linear models and look at residuals?

start with cod
```{r cod_prediction}
# just do non-zero. 
cod_pa <- mfri_pa %>%
  dplyr::select (sample_id, Gadus_morhua) %>%
  rename (Present = Gadus_morhua)

cod_ab <- comb_ds %>%
  filter (species == 1) %>%
  group_by (sample_id) %>%
  summarize (tot_n = sum(n_per_nautmile)) %>%
  mutate (log_ab = log (tot_n)) %>%
  # right_join (cod_pa, by = "sample_id") %>%
  # replace_na (list (tot_n = 0)) %>%
  left_join (mfri_pred, by = "sample_id")

# fit simple linear model of non-zero cod abundance * surface temp
sst_cc <- cod_ab %>%
  filter (!is.na (surface_temp))

M1_sst <- lm (log_ab ~ surface_temp, data = sst_cc)

# model validation from zuur
op <- par (mfrow = c (2,2), mar = c (5,4,1,2))
plot (M1_sst, add.smooth = FALSE, which = 1)
E <- resid (M1_sst)
hist (E, xlab= "Residuals", main = "")
plot (sst_cc$tot_n, E, xlab = "log abundance", ylab = "Residuals")
plot (sst_cc$surface_temp, E, xlab = "SST", ylab = "Residuals")
par (op)
```

try with poisson
```{r cod_poisson}
# Zuur ch 9 pg 216, 9.5.2
plot (sst_cc$surface_temp, sst_cc$tot_n, xlab = "SST", ylab = "tot cod abundance")
P1 <- glm (tot_n ~ surface_temp, family = poisson, data = sst_cc)
summary (P1)
```
