---
title: "megrim notes"
date: "07-23-2020"
output: html_notebook
---
Lepidorhombus_whiffiagonis


Fishbase: 
Depth range 100-700m, usually  100-400
Temp: 6.3-11.4, mean 7.7
25cm maturity, max 60, 

coastal europe and iceland
soft bottoms, small fishes, squids, crustaceans



```{r setup}

library(tidyverse)
library(gridExtra)

library (maps)
library (mapdata)
library (sf)
library (rnaturalearth)
library (rnaturalearthdata)
library (rgeos)
library (viridis)

world <- ne_countries(scale = "medium", returnclass = "sf")

# sample ID of all megrim presences and absences
megrim_pa <- read_csv ("../Data/PA_MFRI.csv",
                     col_types = cols(
                       sample_id = col_factor()
                     )) %>%
  dplyr::select (sample_id, Lepidorhombus_whiffiagonis)

# full MFRI info for all megrim presences, retaining length data
megrim_samp_full <- read_csv ("../Data/MFRI_comb_survey.csv",
                      col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor()
                      )
                      )%>%
  filter (species == 26, !(year == 2011 & season == "autumn")) # remove autumn 2011
  
# all megrim presences, grouped by sample_ID (lengths aggregated
megrim_samp <- megrim_samp_full %>%
  group_by (sample_id) %>%
  summarize (n_tot = sum(n_per_nautmile, na.rm = TRUE),
             kg_tot = sum(kg_per_nautmile, na.rm = TRUE),
             station = first(station),
             lat = first(lat),
             lon = first(lon),
             year = first(year),
             date = first(date),
             season = first(season)
  ) %>%
  mutate (n_log = log (n_tot),
          kg_log = log(kg_tot))


# predictor variables
mfri_pred <- read_csv ("../Data/MFRI_predictor_df.csv",
                       col_types = cols(
                      sample_id = col_factor(),

                      stat_sq = col_factor())
)
```
Present in `r nrow (megrim_samp)` samples (`r nrow(megrim_samp)/29002 * 100`%).

### Histograms

```{r length_season}
megrim_samp_full %>%
  ggplot (aes (y = length_cm, x = season), alpha = 0.5) +
  geom_violin()

megrim_samp_full %>%
  ggplot (aes (x = length_cm, y = n_per_nautmile, fill = season)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.9) 

```
much more common in spring. kind of bimondal, whereas fall has a wider spread. 

## biomass and abundance dotcharts
```{r dotcharts}
dotchart(megrim_samp$n_tot, 
         groups = factor(megrim_samp$season), 
         main = "Abundance")

dotchart(megrim_samp$n_log, 
         #groups = factor(megrim_samp$season), 
         main = "log abundance")

```
one big one over 5, 1989

```{r outlier_check}

megrim_samp %>%
  filter (n_tot > 5)

megrim_samp$n_tot[which (megrim_samp$n_tot > 10)] > quantile (megrim_samp$n_tot, 0.75) + 1.5 * (quantile (megrim_samp$n_tot, 0.75) - quantile (megrim_samp$n_tot, 0.25)) # yes

```


### Trends in abundance and biomass over time
```{r time_series, warning = FALSE}
biomass_ts <- megrim_samp %>%
  #filter (!sample_id == 31881) %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_kg, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Biomass annual mean")+
    theme(legend.position = "none")


ab_ts <- megrim_samp %>%
 # filter (!sample_id == 31881) %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_n, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Abundance annual mean")+
    theme(legend.position = "none")

grid.arrange (biomass_ts, ab_ts, nrow = 1)
```
1989 point does make a spike, but catches high then decline mid-90s, nadir 2000, then rising. 

### Spatial distribution

```{r season_distrib, warning = FALSE}
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = megrim_samp, aes (x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.25, .25)) +
    scale_fill_viridis_c(name = "Mean log abundance") +
    theme_bw() +
    facet_wrap (~season)
```
southwest and south exclusively, no northern


```{r space_by_length, warning = FALSE}
length_bins = quantile (megrim_samp_full$length_cm)
length_bins[1] <- 0 # include smallest ones

megrim_length_bins <- megrim_samp_full %>%
  mutate (length_bin = cut(length_cm, breaks = length_bins),
          n_log = log(n_per_nautmile),
          n_tot = sum(n_per_nautmile))
  
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = megrim_length_bins, aes(x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.5, .5)) +
    scale_fill_viridis_c(name = "Mean log abundance") +
    theme_bw() +
  theme (axis.text = element_blank()) +
    facet_grid (season ~ length_bin)
```
maybe more north in fall?

### Plot abundance vs. predictor variables
```{r megrim_preds, warning = FALSE}
megrim_ab_pred <- megrim_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  right_join (mfri_pred, by = "sample_id")

bt_p <- ggplot (megrim_ab_pred, aes (x = bottom_temp, y = kg_log)) +
  geom_point(alpha = 0.5) +
  geom_smooth (method = "lm") +
  ggtitle ("Bottom temp")
  

sst_p <- ggplot (megrim_ab_pred, aes (x = surface_temp, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Surface temp")

sal_p <- ggplot (megrim_ab_pred, aes (x = gins_sal, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Salinity")

lat_p <- ggplot (megrim_ab_pred, aes (x = lat, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Latitude")

lon_p <- ggplot (megrim_ab_pred, aes (x = lon, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Longitude")

depth_p <-  ggplot (megrim_ab_pred, aes (x = tow_depth_begin, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Depth")


grid.arrange (sst_p, bt_p, sal_p, lat_p, lon_p, depth_p, nrow = 2)

```
up siwht all env and depth. strong down with latitude, slightly up longitude

### Model results
```{r model_explore}
library (mgcv)
load ("../Models/PresAbs/Lepidorhombus_whiffiagonis_PA_full.Rdata")
summary (gam_PA)

```


Partial plots
```{r partial_plots_PA, warning = FALSE}

par (mfrow = c (2, 3), mar = c (2, 1.5, 1.5, 1.5), oma = c(0,1,2,0))
plot (gam_PA, residuals = TRUE, 
      select = 1, 
      ylab = "", xlab = "", 
      main = paste ("te(Lon, Lat)", round (summary (gam_PA)$edf[1], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 2, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sst)", round (summary (gam_PA)$edf[2], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 3, 
      ylim = c (-10, 10), 
      ylab = "", xlab = "", 
      main = paste ("s(bt)", round (summary (gam_PA)$edf[3], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 4, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sal)", round (summary (gam_PA)$edf[4], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 5, 
      ylim = c (-10, 25), 
      ylab = "", xlab = "", 
      main = paste ("s(depth)", round (summary (gam_PA)$edf[5], 2), sep = ", "))


```
sal didn't converge

```{r single_Var_drop_PA}
full.terms <- c ("te (lon, lat)", "year", "season", "s(surface_temp)", "s(bottom_temp)", "s(gins_sal)", "s(tow_depth_begin)")

megrim_pred_PA <- megrim_pa %>%
  left_join (mfri_pred, by = "sample_id") %>%
  rename ("Presence" = Lepidorhombus_whiffiagonis)

single.var.comp <- data.frame()
for (x in full.terms){
  # single variable model
  formula <- as.formula (paste0 ("Presence ~", x))
  G <- gam (formula, family = binomial, data = megrim_pred_PA)
  
  # full model without that variable
 formula_drop <- as.formula (paste0 ("Presence ~", (paste (full.terms[! full.terms %in% x], collapse = " + "))))
 G_drop <- gam (formula_drop, family = binomial, data = filter (megrim_pred_PA, !is.na (x)))
 

  # extract summaries, put together in temporary df
  df <- data.frame (Var = x, 
                    R2_full = round (summary (gam_PA)$dev.expl * 100, 2), # gam_PA is full model
                
                    R2_s = round (summary(G)$dev.expl * 100, digits = 2),
                   
                    R2_d =  round (summary(G_drop)$dev.expl * 100, digits = 2),
                    #N_full = length (G_full$y),
                    
                    Diff_full = round (summary(G_drop)$dev.expl * 100 - summary (gam_PA)$dev.expl * 100, 2)
  )
  
  # rbind to full data frame
  single.var.comp <- rbind (single.var.comp, df)
  
}


single.var.comp

save (single.var.comp, file = "../Models/PresAbs_summary_tables/megrim_single_var_comp.RData")

```
everything contributes. mostly lat and lon. depth has much smaller r2 single than temps or salinity, but biggest drop diff

```{r biomass_gam}

load ("../Models/Biomass/Lepidorhombus_whiffiagonis_LB_full.Rdata")
summary (gam_LB)
```
year not significant. 

```{r partial_plots_lb, warning = FALSE}

par (mfrow = c (2, 3), mar = c (2, 1.5, 1.5, 1.5), oma = c(0,1,2,0))
plot (gam_LB, residuals = TRUE, 
      select = 1, 
      ylab = "", xlab = "", 
      main = paste ("te(Lon, Lat)", round (summary (gam_LB)$edf[1], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 2, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sst)", round (summary (gam_LB)$edf[2], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 3, 
      ylim = c (-10, 10), 
      ylab = "", xlab = "", 
      main = paste ("s(bt)", round (summary (gam_LB)$edf[3], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 4, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sal)", round (summary (gam_LB)$edf[4], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 5, 
      #ylim = c (-10, 15), 
      ylab = "", xlab = "", 
      main = paste ("s(depth)", round (summary (gam_LB)$edf[5], 2), sep = ", "))

```


```{r single_Var_drop_lb}
#full.terms <- c ("te (lon, lat)", "year", "season", "s(surface_temp)", "s(bottom_temp)", "s(gins_sal)", "s(tow_depth_begin)")

single.var.comp <- data.frame()
for (x in full.terms){
  # single variable model
  formula <- as.formula (paste0 ("kg_log ~", x))
  G <- gam (formula, family = gaussian, data = megrim_ab_pred)
  
  # full model without that variable
 formula_drop <- as.formula (paste0 ("kg_log ~", (paste (full.terms[! full.terms %in% x], collapse = " + "))))
 G_drop <- gam (formula_drop, family = gaussian, data = filter (megrim_ab_pred, !is.na (x)))
 

  # extract summaries, put together in temporary df
  df <- data.frame (Var = x, 
                    R2_full = round (summary (gam_LB)$dev.expl * 100, 2), # gam_PA is full model
                
                    R2_s = round (summary(G)$dev.expl * 100, digits = 2),
                   
                    R2_d =  round (summary(G_drop)$dev.expl * 100, digits = 2),
                    #N_full = length (G_full$y),
                    
                    Diff_full = round (summary(G_drop)$dev.expl * 100 - summary (gam_LB)$dev.expl * 100, 2)
  )
  
  # rbind to full data frame
  single.var.comp <- rbind (single.var.comp, df)
  
}


single.var.comp

save (single.var.comp, file = "../Models/Biomass_summary_tables/megrim_single_var_comp.RData")

```
year not helpful. bottom temp makes worse. 