---
title: "Tusk notes"
date: "07-15-2020"
output: html_notebook
---
Brosme_brosme

brosme comes from breek bibrosko, to devour


Fishbase: 
Depth range 18-1000m, usually 18-549 (esp for northeast atlantic). 
Temp: 0.5-7.7
53 cm at maturity, max 120, max 30kg

Northwest atlantic, northeast atlantic, "off iceland"
rough rock, gravel bottoms, far from the shore, near the bottom. 
crustaceans and shellfishes, benthic fishes, starfishes. predators are seals. 
spawning grounds btw scotland and iceland, 200-500m



```{r setup}

library(tidyverse)
library(gridExtra)

library (maps)
library (mapdata)
library (sf)
library (rnaturalearth)
library (rnaturalearthdata)
library (rgeos)
library (viridis)

world <- ne_countries(scale = "medium", returnclass = "sf")

# sample ID of all tusk presences and absences
tusk_pa <- read_csv ("../Data/PA_MFRI.csv",
                     col_types = cols(
                       sample_id = col_factor()
                     )) %>%
  select (sample_id, Brosme_brosme)

# full MFRI info for all tusk presences, retaining length data
tusk_samp_full <- read_csv ("../Data/MFRI_comb_survey.csv",
                      col_types = cols(
                      sample_id = col_factor(),
                      species = col_factor(),
                      stat_sq = col_factor()
                      )
                      )%>%
  filter (species == 8, !(year == 2011 & season == "autumn")) # remove autumn 2011
  
# all tusk presences, grouped by sample_ID (lengths aggregated
tusk_samp <- tusk_samp_full %>%
  group_by (sample_id) %>%
  summarize (n_tot = sum(n_per_nautmile, na.rm = TRUE),
             kg_tot = sum(kg_per_nautmile, na.rm = TRUE),
             station = first(station),
             lat = first(lat),
             lon = first(lon),
             year = first(year),
             date = first(date),
             season = first(season)
  ) %>%
  mutate (n_log = log (n_tot),
          kg_log = log(kg_tot))


# predictor variables
mfri_pred <- read_csv ("../Data/MFRI_predictor_df.csv",
                       col_types = cols(
                      sample_id = col_factor(),

                      stat_sq = col_factor())
)
```
Present in `r nrow (tusk_samp)` samples (`r nrow(tusk_samp)/29002 * 100`%). Middling

### Histograms

```{r length_season}
tusk_samp_full %>%
  ggplot (aes (y = length_cm, x = season), alpha = 0.5) +
  geom_violin()

tusk_samp_full %>%
  ggplot (aes (x = length_cm, y = n_per_nautmile, fill = season)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.9) 

```
Similar spread but more small individuals in spring. 

## biomass and abundance dotcharts
```{r dotcharts}
dotchart(tusk_samp$n_tot, 
         groups = factor(tusk_samp$season), 
         main = "Abundance")

dotchart(tusk_samp$n_log, 
         #groups = factor(tusk_samp$season), 
         main = "log abundance")

```
A handful above 2, 3 in 2006, probably same square--281092, 281080, 281093. actually stat square is in station id, but 361 stands out, 4 of the 6 are there. 
```{r outlier_check}

tusk_samp %>%
  filter (n_tot > 2)

tusk_samp$n_tot[which (tusk_samp$n_tot > 2)] > quantile (tusk_samp$n_tot, 0.75) + 1.5 * (quantile (tusk_samp$n_tot, 0.75) - quantile (tusk_samp$n_tot, 0.25)) # yes

```



### Trends in abundance and biomass over time
```{r time_series, warning = FALSE}
biomass_ts <- tusk_samp %>%
  #filter (!sample_id == 31881) %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_kg, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Biomass annual mean")+
    theme(legend.position = "none")


ab_ts <- tusk_samp %>%
 # filter (!sample_id == 31881) %>%
  group_by (year, season) %>%
  summarize (mn_kg = mean (kg_tot),
             se_kg = var(kg_tot)/sqrt (n()),
             mn_n = mean(n_tot),
             se_n = var (n_tot)/sqrt(n())
  ) %>%
  ggplot (aes (x = year, y = mn_n, col = season)) +
    geom_point() +
    geom_line() +
    #geom_pointrange (aes (ymin = mn_kg - se_kg, ymax = mn_kg + se_kg)) +
    geom_smooth (method = "lm") +
    ggtitle("Abundance annual mean")+
    theme(legend.position = "none")

grid.arrange (biomass_ts, ab_ts, nrow = 1)
```
Biomass and abundance look pretty different, maybe they've been getting bigger in the fall? or big spike in abundance 2004-2012ish was mostly small spring individuals? lower catch 1990-2003ish seems shared.  

### Spatial distribution

```{r season_distrib, warning = FALSE}
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = tusk_samp, aes (x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.25, .25)) +
    scale_fill_viridis_c(name = "Mean log abundance") +
    theme_bw() +
    facet_wrap (~season)
```
Further offshore, especially in autumn. concentrated in southeast and reykjanes in spring. hotspot spring looks like square 361 


```{r space_by_length, warning = FALSE}
length_bins = quantile (tusk_samp_full$length_cm)
length_bins[1] <- 0 # include smallest ones

tusk_length_bins <- tusk_samp_full %>%
  mutate (length_bin = cut(length_cm, breaks = length_bins),
          n_log = log(n_per_nautmile),
          n_tot = sum(n_per_nautmile))
  
ggplot(data = world) +
  geom_sf() +
  coord_sf (xlim = c(-30, -8), ylim = c (62,70), expand = FALSE ) +
  stat_summary_2d(data = tusk_length_bins, aes(x = lon, y = lat, z = n_log), fun = mean, binwidth = c (.5, .5)) +
    scale_fill_viridis_c(name = "Mean log abundance") +
    theme_bw() +
  theme (axis.text = element_blank()) +
    facet_grid (season ~ length_bin)
```
Not sure I see a pattern here. possibly larger move further offshore in autumn, but might just be that there are fewer small individuals in autumn. 

### Plot abundance vs. predictor variables
```{r tusk_preds, warning = FALSE}
tusk_ab_pred <- tusk_samp %>%
  dplyr::select (sample_id, n_log, kg_log) %>%
  right_join (mfri_pred, by = "sample_id")

bt_p <- ggplot (tusk_ab_pred, aes (x = bottom_temp, y = kg_log)) +
  geom_point(alpha = 0.5) +
  geom_smooth (method = "lm") +
  ggtitle ("Bottom temp")
  

sst_p <- ggplot (tusk_ab_pred, aes (x = surface_temp, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Surface temp")

sal_p <- ggplot (tusk_ab_pred, aes (x = gins_sal, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Salinity")

lat_p <- ggplot (tusk_ab_pred, aes (x = lat, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Latitude")

lon_p <- ggplot (tusk_ab_pred, aes (x = lon, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Longitude")

depth_p <-  ggplot (tusk_ab_pred, aes (x = tow_depth_begin, y = kg_log)) +
  geom_point() +
  geom_smooth (method = "lm")+
  ggtitle ("Depth")


grid.arrange (sst_p, bt_p, sal_p, lat_p, lon_p, depth_p, nrow = 2)

```
Pos with temp, sal, depth. negative with latitude and longitude, similar to blue ling. would have expected different lat/lon pattern with southwest cluster, but maybe lon trend isn't that strong or nonlinear. 

### Model results
```{r model_explore}
library (mgcv)
load ("../Models/PresAbs/Brosme_brosme_PA_full.Rdata")
summary (gam_PA)

```
not so great, only 30%. year not significant. 

Partial plots
```{r partial_plots_PA, warning = FALSE}

par (mfrow = c (2, 3), mar = c (2, 1.5, 1.5, 1.5), oma = c(0,1,2,0))
plot (gam_PA, residuals = TRUE, 
      select = 1, 
      ylab = "", xlab = "", 
      main = paste ("te(Lon, Lat)", round (summary (gam_PA)$edf[1], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 2, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sst)", round (summary (gam_PA)$edf[2], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 3, 
      ylim = c (-10, 10), 
      ylab = "", xlab = "", 
      main = paste ("s(bt)", round (summary (gam_PA)$edf[3], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 4, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sal)", round (summary (gam_PA)$edf[4], 2), sep = ", "))
plot (gam_PA, residuals = TRUE, 
      select = 5, 
      ylim = c (-10, 25), 
      ylab = "", xlab = "", 
      main = paste ("s(depth)", round (summary (gam_PA)$edf[5], 2), sep = ", "))


```
really weird depth. bottom temp optimal 2-7ish. 

```{r single_Var_drop_PA}
full.terms <- c ("te (lon, lat)", "year", "season", "s(surface_temp)", "s(bottom_temp)", "s(gins_sal)", "s(tow_depth_begin)")

tusk_pred_PA <- tusk_pa %>%
  left_join (mfri_pred, by = "sample_id") %>%
  rename ("Presence" = Brosme_brosme)

single.var.comp <- data.frame()
for (x in full.terms){
  # single variable model
  formula <- as.formula (paste0 ("Presence ~", x))
  G <- gam (formula, family = binomial, data = tusk_pred_PA)
  
  # full model without that variable
 formula_drop <- as.formula (paste0 ("Presence ~", (paste (full.terms[! full.terms %in% x], collapse = " + "))))
 G_drop <- gam (formula_drop, family = binomial, data = filter (tusk_pred_PA, !is.na (x)))
 

  # extract summaries, put together in temporary df
  df <- data.frame (Var = x, 
                    R2_full = round (summary (gam_PA)$dev.expl * 100, 2), # gam_PA is full model
                
                    R2_s = round (summary(G)$dev.expl * 100, digits = 2),
                   
                    R2_d =  round (summary(G_drop)$dev.expl * 100, digits = 2),
                    #N_full = length (G_full$y),
                    
                    Diff_full = round (summary(G_drop)$dev.expl * 100 - summary (gam_PA)$dev.expl * 100, 2)
  )
  
  # rbind to full data frame
  single.var.comp <- rbind (single.var.comp, df)
  
}


single.var.comp

save (single.var.comp, file = "../Models/PresAbs_summary_tables/tusk_single_var_comp.RData")

```

Year makes model worse. lat/lon doing the most here, and depth. sst not adding much. 

```{r biomass_gam}

load ("../Models/Biomass/Brosme_brosme_LB_full.Rdata")
summary (gam_LB)
```
bottom temp not significant. very low dev explained.  

```{r partial_plots_lb, warning = FALSE}

par (mfrow = c (2, 3), mar = c (2, 1.5, 1.5, 1.5), oma = c(0,1,2,0))
plot (gam_LB, residuals = TRUE, 
      select = 1, 
      ylab = "", xlab = "", 
      main = paste ("te(Lon, Lat)", round (summary (gam_LB)$edf[1], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 2, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sst)", round (summary (gam_LB)$edf[2], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 3, 
      ylim = c (-10, 10), 
      ylab = "", xlab = "", 
      main = paste ("s(bt)", round (summary (gam_LB)$edf[3], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 4, 
      ylim = c (-5, 5), 
      ylab = "", xlab = "", 
      main = paste ("s(sal)", round (summary (gam_LB)$edf[4], 2), sep = ", "))
plot (gam_LB, residuals = TRUE, 
      select = 5, 
      #ylim = c (-10, 15), 
      ylab = "", xlab = "", 
      main = paste ("s(depth)", round (summary (gam_LB)$edf[5], 2), sep = ", "))

```
higher at 6-10 sst

```{r single_Var_drop_lb}
#full.terms <- c ("te (lon, lat)", "year", "season", "s(surface_temp)", "s(bottom_temp)", "s(gins_sal)", "s(tow_depth_begin)")

single.var.comp <- data.frame()
for (x in full.terms){
  # single variable model
  formula <- as.formula (paste0 ("kg_log ~", x))
  G <- gam (formula, family = gaussian, data = tusk_ab_pred)
  
  # full model without that variable
 formula_drop <- as.formula (paste0 ("kg_log ~", (paste (full.terms[! full.terms %in% x], collapse = " + "))))
 G_drop <- gam (formula_drop, family = gaussian, data = filter (tusk_ab_pred, !is.na (x)))
 

  # extract summaries, put together in temporary df
  df <- data.frame (Var = x, 
                    R2_full = round (summary (gam_LB)$dev.expl * 100, 2), # gam_PA is full model
                
                    R2_s = round (summary(G)$dev.expl * 100, digits = 2),
                   
                    R2_d =  round (summary(G_drop)$dev.expl * 100, digits = 2),
                    #N_full = length (G_full$y),
                    
                    Diff_full = round (summary(G_drop)$dev.expl * 100 - summary (gam_LB)$dev.expl * 100, 2)
  )
  
  # rbind to full data frame
  single.var.comp <- rbind (single.var.comp, df)
  
}


single.var.comp

save (single.var.comp, file = "../Models/Biomass_summary_tables/tusk_single_var_comp.RData")

```
bottom temp makes worse. lat/lon the only thing really doing anything here, then maybe depth.  
